---
title: "Re-routing Harbor Seal Movement Tracks Around Land with {pathroutr}"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Re-routing Harbor Seal Movement Tracks Around Land with {pathroutr}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Alaska Harbor Seal Example

```{r setup}
library(pathroutr)
```

## Essential Packages

This example relies on two non-release packages of `{stplanr}` and `{crawl}`

```{r eval=FALSE}
# not run
library(remotes)
install_github('NMML/crawl@devel')
install_github('ropensci/stplanr')
```

```{r}
library(crawl)
library(stplanr)
```

Additional packages we'll want to load

```{r}
library(dplyr)
library(sf)
library(ggplot2)
library(ggspatial)
library(dplyr)
```

The purpose of this is to demonstrate a more 'real world' use of `{pathroutr}` for
re-routing of marine animal movement around land features. Here, we will use the
Alaska harbor seal data provided in the `{crawl}` package to:

1. demonstrate and data wrangling and pre-processing needed
2. use `{crawl`}'s movement model to predict the most likely path
3. re-route the predicted path around any land barriers
4. repeat the process but using `{crawl}`'s multiple imputation functionality to 
create a set of possible predicted paths that span model uncertainty

## Pre-processing Track Data and Land Polygon Data

For our land data, we will source in the Alaska 1:250000 coastal data polygon. This
is provided by the Alaska Department of Natural Resources and was obtained from their
open data portal (https://gis.data.alaska.gov/datasets/alaska-1250000). The commented
code provides a query to 
pull the data directly from the portal API. Note, only those polygons that intersect
with the bounding box of our harbor seal movement are included.

NOTE: because this step is time consuming, we will load directly from package data

```{r}
# akcoast_qry <- "https://arcgis.dnr.alaska.gov/arcgis/rest/services/OpenData/Physical_AlaskaCoast/MapServer/2/query?where=1%3D1&outFields=*&geometry=-159.240%2C55.112%2C-152.422%2C59.413&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=3338&f=json"
# 
# akcoast <- sf::read_sf(akcoast_qry)
# akcoast <- sf::st_make_valid(akcoast)
data("akcoast")
```

Now, let's load in our harbor seal data from the `{crawl}` package. The `harborSeal_sf`
data is only available in the *devel* branch. We also transform to EPSG:3338.

```{r}
data("harborSeal_sf")

harborSeal_sf <- harborSeal_sf %>% sf::st_transform(3338)
```

Let's plot things to make sure everything looks good. The `harborSeal_sf` data has
missing values in the *geometry* field so we need to filter those out when making a line.

```{r}
l <- harborSeal_sf %>% 
  dplyr::filter(!sf::st_is_empty(.)) %>% 
  dplyr::summarise(do_union = FALSE) %>% 
  sf::st_cast('LINESTRING')

ggplot() +
  ggspatial::annotation_spatial(akcoast, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(l, color = "darkgrey", size = 0.5) +
  ggspatial::layer_spatial(harborSeal_sf, color = "deepskyblue3", size = 0.5) +
  theme_void() +
  ggtitle("Raw Argos Ovservations and Derived Trackline",
          subtitle = "Harbor Seal, Alaska, USA")
```

## Re-Route Raw Observations

While the preferred approach would be to rely on a movement model for predicting the
path of our seal, it can sometimes be useful to just correct the raw observations. So,
that's what we'll do first.

### Limit Barrier Polygon to Region of Interest

With such a large geographic area, the computational time needed to create our
visibility graph could be quite large. So, we will want to limit our land polygon
as much as we reasonably can.

We'll do this by creating a convex hull boundary around our observed data and limiting
the region to this space. Here, we'll go with a buffer of 35 km around each point and
rely on the buffered features to create our convex hull.

```{r}
land_region <- sf::st_buffer(harborSeal_sf, dist = 35000) %>% 
  sf::st_union() %>% 
  sf::st_convex_hull() %>% 
  sf::st_intersection(akcoast) %>% 
  st_collection_extract('POLYGON') %>% 
  st_sf()

ggplot() +
  ggspatial::annotation_spatial(land_region, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(l, color = "darkgrey", size = 0.5) +
  ggspatial::layer_spatial(harborSeal_sf, color = "deepskyblue3", size = 0.5) +
  theme_void() +
  ggtitle("Land Barrier Polygon",
          subtitle = "limited to 35km convex hull around track points")
```

### Create Visibility Graph

At this point, we are ready to create our visibility graph using the 
`pathroutr::prt_visgraph()` function. We'll keep *centroids = FALSE* in order to 
speed up the build. Adding additional centroids will increase the likelihood of
linearity in the calculated shortest path. This is especially true when the shortest
path crosses larger areas of open water. However, adding the additional centroids
results in a very large increase in computational time (both in createion of the visual
graph and in subesquent operations). There is likely little improvement in
situations where the shortest path routing follows the coast. The default setting
is *centroids=FALSE*. If a user chooses to set *centroids=TRUE*, we strongly 
encourage setting *centroid_limit* to a large value (default is *1e+07*).

NOTE: the following step will take approximately 30 minutes. Here, to save time, the
code is not evaluated and we will load from package data.

```{r, eval=FALSE}
vis_graph <- prt_visgraph(land_region)
```

Let's take a look at our network. The 'edges' are stored within `@sl` slot of `vis_graph`.
We can pass thing along to `ggspatial::layer_spatial()` just like an `sf` object.

```{r}
data("vis_graph")
ggplot() +
  ggspatial::annotation_spatial(land_region, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(vis_graph@sl, size = 0.5) +
  theme_void()+
  ggtitle("Visibility Graph Edges")
```

Note the large triangles in the large, open water areas. This is a key advantage for our
purpose as we concentrate the detail of our network along the coastline where routing
of our tracks around land requires the most detail. The trade-off is that any shortest 
path routing that crosses the large, open areas will have less linearity. If we had set
*centroids=FALSE*, then a centroid point would have been added to each of these triangles
and the mesh would have been rebuild to include those points.

### Determine Segments of Track On Land

In this example, our interest is to re-route only the portions of the track that cross 
land. So, we need to identify the segments of consecutive points on land and, also, the
points in water that bookend each segment. We will use this information to calculate
our shortest path through our network and create an updated series of point features.

Since our bookend points were not used to create our visibility graph, we need to also
determine the closes network node to each point.

```{r}
track_pts <- harborSeal_sf %>% dplyr::filter(!sf::st_is_empty(.))
segs_tbl <- get_barrier_segments(track_pts,land_region)
segs_tbl <- segs_tbl %>% prt_nearestnode(vis_graph, maxdist = 50000)
segs_tbl
```

### Determine Re-routed Track via Shortest Path

Now, we have all the information we need to calculate the shortest path through our
network for each segment. The `stplanr::sum_network_routes()` function only calculates
the LINESTRING from start node to end node. So, the `prt_shortpath()` function extends
each end to connect with our bookend points for each segment.

```{r}
segs_tbl <- segs_tbl %>% prt_shortpath(vis_graph)

ggplot() + 
  ggspatial::annotation_spatial(land_region, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(segs_tbl$geometry, color = "deepskyblue3") +
  theme_void() +
  ggtitle("Re-routed Segments That Were Identified to Cross Land")
```

### Update Geometry With Re-routed Coordinates

This last bit is the point where we insert the fixed points back into our original
path.

```{r}
track_pts_fix <- prt_update_points(track_pts, segs_tbl)
```

```{r}
track_line_fixed <- track_pts_fix %>% summarise(do_union = FALSE) %>% st_cast('LINESTRING')

ggplot() +
  ggspatial::annotation_spatial(land_region, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(track_line_fixed, color = "darkgrey", size = 0.5) +
  ggspatial::layer_spatial(track_pts_fix, color = "deepskyblue3", size = 0.5) +
  theme_void() +
  ggtitle("Trackline and Argos Observations After Re-routing")
```

## Correcting a Movement Model Prediction

First, let's fit our movement model. Here, we rely on an example from the `{crawl}`
package. But, other model approaches (e.g. `{foieGras}`) should work as well. We have
plans to provide specific methods for a variety of model outputs.

### Fit Movement Model and Predict Locations

As you might have noticed when we re-routed our raw observations, the trackline derived
from those points can still cross land. This is because we only control for the presence
of observation points on land. At the irregular and sometimes large time intervals
common to location data for marine mammals, this cannot be avoided. However, once we
have a movement model, we can predicted a finer location intervals and limit this
possibility. You will notice in this example, the prediction interval is every 5 minutes.

```{r}
##Fit model as given in Johnson et al. (2008) Ecology 89:1208-1215
## Start values for theta come from the estimates in Johnson et al. (2008)
fixPar = c(log(250), log(500), log(1500), rep(NA,5), 0)
displayPar( mov.model=~1, err.model=list(x=~Argos_loc_class-1),data=harborSeal_sf, 
            activity=~I(1-DryTime),fixPar=fixPar)
constr=list(
  lower=c(rep(log(1500),3), rep(-Inf,2)),
  upper=rep(Inf,5)
)

set.seed(123)
fit1 <- crwMLE(
  mov.model=~1, err.model=list(x=~Argos_loc_class-1), activity=~I(1-DryTime),
  data=harborSeal_sf,  Time.name="Time", 
  fixPar=fixPar, theta=c(rep(log(5000),3),log(3*3600), 0),
  constr=constr, method="L-BFGS-B",
  control=list(maxit=2000, trace=1, REPORT=1)
)

pred1 = crwPredict(fit1, predTime = '5 min')
pred1_sf <- pred1 %>% crw_as_sf("POINT","p")
```

### Identify Predicted Locations on Land

As before, we need to identify all of the segments of consecutive locations on land and
determine the associated start and end nodes within our network.

```{r}
segs_tbl <- get_barrier_segments(pred1_sf,land_region)
segs_tbl <- segs_tbl %>% prt_nearestnode(vis_graph)
segs_tbl
```

### Re-route Model Prediction via Shortest Path

As before, we use the `prt_shortpath()` function to calculated all of our re-routed paths

```{r}
segs_tbl <- segs_tbl %>% prt_shortpath(vis_graph)

ggplot() + 
  ggspatial::annotation_spatial(land_region, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(segs_tbl$geometry, color = "deepskyblue3") +
  theme_void() +
  ggtitle("Re-routed Segments from Predicted Path")
```

### Update Predicted Path

This last bit is the point where we insert the fixed points back into our original
path. This is done by taking the number of points in the original segment and 
distributing an equal number of points along the updated path. At this time, the geometry 
is simply updated in place. This isn't, technically, the correct approach as all of the 
model prediction parameters are no longer valid for these updated records. Expect this
workflow/function/approach to change. But, for now, it allows us to demonstrate the
issue and troubleshoot any issues.

```{r}
track_pts_fix <- prt_update_points(pred1_sf, segs_tbl)
```

```{r}
track_line_fixed <- track_pts_fix %>% summarise(do_union = FALSE) %>% st_cast('LINESTRING')

ggplot() +
  ggspatial::annotation_spatial(land_region, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(track_line_fixed, color = "deepskyblue3", size = 0.5) + 
  theme_void() +
  ggtitle("Updated Predicted Path to Re-route Around Land",
          subtitle = "Harbor Seal, Alaska")
```
