---
title: "Re-routing Straight Line Paths Around Barriers - A Visibility Graph Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Re-routing Straight Line Paths Around Barriers - A Visibility Graph Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the basic concept within the `{pathroutr}` package for
re-routing paths that cross a barrier around the barrier using the shortest path
through a visibility graph.

The larger intent is to use this approach for adjust the movement tracks of marine
animals through water when estimated paths incorrectly cross land.

```{r setup}
library(pathroutr)
data(land_barrier)
data(poi)
```

## The Test Environment

```{r}
library(dplyr)
library(ggplot2)
library(sf)
library(ggspatial)
```


First, let's setup our test space with our included land barrier and points of interest.
This was based off of the north Kodiak Island region in Alaska, USA (but, is not an
exact replica. So, not for navigational use!!). 

```{r}
ggplot() + 
  ggspatial::annotation_spatial(data = land_barrier, 
                                fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(data = poi) +
  theme_void()
```

The complexity of the nearshore environment with small islands and narrow passages is
a difficult situation for estimating marine animal tracks from telemetry devices that
often have error associated with the location estimates.

Here, we sample 15 of our points of interest and, then, connect them as a line
representing a path across our complicated landscape.

```{r}
l_pts <- poi %>% slice_sample(n=15)
path <- l_pts %>% summarise(do_union = FALSE) %>% st_cast('LINESTRING')

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(poi) +
  ggspatial::layer_spatial(path, color = "deepskyblue3") +
  ggspatial::layer_spatial(l_pts[1,], color = "darkgreen", size = 4) +
  ggspatial::layer_spatial(l_pts[15,], color = "darkred", size = 4) +
  theme_void()
```

Now, let's sample 10000 points along this string and this will represent our simulated
path of track observations.

```{r}
track_pts <- st_sample(path, size = 10000, type = "regular")

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(path, color = "deepskyblue3") +
  ggspatial::layer_spatial(track_pts) +
  theme_void()

```

## Re-route the path

The first function we'll use form the `{pathroutr}` package is `get_barrier_segments()`
which identifies all of the consecutive track points that intersect with the land
barrier. 

```{r}
segs_tbl <- get_barrier_segments(track_pts,land_barrier)
segs_tbl
```

Our next step is to create a visibility graph. This is, essentially, our road network for
our marine environment. At some point, I'll provide greater detail on this process. But,
in simple terms, we connect all of the vertices for our barrier polygon with a Delaunay
triangle mesh and, then, remove any of the edges that cross land. Our `prt_visgraph()`
function returns a _SpatialLinesNetwork_ created with the `{stplanr}` package.

With our network created, we need to identify the nearest node on the network for 
the start and end points of each identified segment.

NOTE: there is currently an error when the nearest node for the start and end is the same
node when calculating the shortest path between them. I need to explore this further.

```{r}
vis_graph <- prt_visgraph(land_barrier, buffer = 150)
segs_tbl <- segs_tbl %>% prt_nearestnode(vis_graph)
```

Calculate all of our shortest path routes around the barrier for each segment

```{r}
segs_tbl <- segs_tbl %>% prt_shortpath(vis_graph)

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(segs_tbl$geometry, color = "deepskyblue3") +
  theme_void()

```

This last bit is the point where we insert the fixed points back into our original
path. It's currently a bit messy and I'll soon turn this into a standalone function.

```{r}
track_pts <- track_pts %>% st_cast('POINT') %>% 
  st_sf() %>% tibble::rowid_to_column("id")

pts_list <- vector(mode = "list", length = nrow(segs_tbl))

for (i in 1:nrow(segs_tbl)) {
  fix_idx <- seq(segs_tbl$start_idx[i]+1,segs_tbl$end_idx[i]-1)
  fix_pts <- sf::st_sample(segs_tbl$geometry[i], size = segs_tbl$n_pts[i], 
                           type = "regular") %>% 
    sf::st_cast('POINT')
  fix_pts <- prt_orient_line(segs_tbl$start_pt[i], fix_pts)
  pts_list[[i]] <- sf::st_sf(id = fix_idx, geometry = fix_pts )
}

fixed_pts <- do.call(rbind, pts_list) %>% 
  tibble::as_tibble()

track_pts_fix <- track_pts %>% tibble::as_tibble()  %>% 
  dplyr::rows_update(fixed_pts, by = "id") %>% 
  sf::st_sf(sf_column_name = "geometry")

fixed_line <- track_pts_fix %>% 
  dplyr::summarise(do_union = FALSE) %>% 
  st_cast('LINESTRING')
```


## Did it work?

A final plot of our corrected path. The green dot is the start and the red dot then end.
The blue line is the shortest path fix and the black line is the new path. Red line
represents the original path.

If everything is working correctly, there should not be any black lines crossing land.

```{r}
track_pts <- track_pts %>% summarise(do_union=FALSE) %>% st_cast('LINESTRING')

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(track_pts, color = "red3") +
  ggspatial::layer_spatial(segs_tbl$geometry, color = "deepskyblue3", size=2) +
  ggspatial::layer_spatial(fixed_line) +
  ggspatial::layer_spatial(l_pts[1,], color = "darkgreen", size = 4) +
  ggspatial::layer_spatial(l_pts[15,], color = "darkred", size = 4) +
  theme_void()
```
