---
title: "Re-routing Straight Line Paths Around Barriers - A Visibility Graph Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Re-routing Straight Line Paths Around Barriers - A Visibility Graph Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the basic concept within the `{pathroutr}` package for
re-routing paths that cross a barrier around the barrier using the shortest path
through a visibility graph.

The larger intent is to use this approach for adjusting the movement tracks of marine
animals through water when estimated paths incorrectly cross land.

```{r setup}
library(pathroutr)
data(land_barrier)
data(poi)
```

## The Test Environment

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(sf)
library(ggspatial)
```


First, let's setup our test space with our included land barrier and points of interest.
This was based off of the north Kodiak Island region in Alaska, USA (but, is not an
exact replica. So, not for navigational use!!). 

```{r}
ggplot() + 
  ggspatial::annotation_spatial(data = land_barrier, 
                                fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(data = poi) +
  theme_void()
```

The complexity of the nearshore environment with small islands and narrow passages is
a difficult situation for estimating marine animal tracks from telemetry devices that
often have error associated with the location estimates.

Here, we sample 15 of our points of interest and, then, connect them as a line
representing a path across our complicated landscape.

```{r}
l_pts <- poi %>% slice_sample(n = 15)
path <- l_pts %>% summarise(do_union = FALSE) %>% st_cast('LINESTRING')

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(poi) +
  ggspatial::layer_spatial(path, color = "deepskyblue3") +
  ggspatial::layer_spatial(l_pts[1,], color = "darkgreen", size = 4) +
  ggspatial::layer_spatial(l_pts[15,], color = "darkred", size = 4) +
  theme_void()
```

Now, let's sample 10000 points along this string and this will represent our simulated
path of track observations.

```{r}
track_pts <- st_sample(path, size = 10000, type = "regular")

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(path, color = "deepskyblue3") +
  ggspatial::layer_spatial(track_pts) +
  theme_void()

```

## Re-route the path

The first function we'll use fromm the `{pathroutr}` package is `get_barrier_segments()`
which identifies all of the consecutive track points that intersect with the land
barrier. The result of this function is a tibble that stores the key metadata about
each of the segments identified to cross land. The *start_idx* and *end_idx* columns
store the row index of the two points that bookend each stretch of consecutive points
that intersect with the barrier polygon. These points DO NOT intersect with the barrier.
The *n_pts* column is the number of consecutive points that intersect with the barrier.
Lastly, the *start_pt* and *end_pt* columns store the Simple Feature geometry for the
bookend points represented by *start_idx* and *end_idx*. Later, these geometries will be
used to identify the nearest node in our network for shortest path routing and the final, 
updated path will start and end with these geometries.

```{r}
segs_tbl <- get_barrier_segments(track_pts,land_barrier)
segs_tbl
```

Our next step is to create a visibility graph. This is, essentially, a road network for
our marine environment. In simple terms, we connect all of the vertices for our barrier 
polygon with a Delaunay triangle mesh and, then, remove any of the edges that cross land. 
Our `prt_visgraph()` function returns a _SpatialLinesNetwork_ created with the `{stplanr}` 
package.

The most computationally intensive step is likely to be identification of which edges
intersect with the barrier polygon. We try to minimize this by, first, buffering the 
barrier by -1 meter and then calling `sf::st_intersect()`. Because all of the nodes in our
network are also vertices in our barrier polygon, all edges intersect with the barrier
polygon. By buffering by -1 meter, we disconnect the nodes. If we didn't do this, we would
need to evaluate two spatial predicate functions (`sf::st_crosses()` and `sf::st_within()`).
With the negative buffer adjustment, we can rely solely on `sf::st_intersect()`. The other
step we do to minimize compute time is to NOT evaluate which edges intersect with the
barrier polygon feature. Instead we evaluate which polygon features intersect with which
edges. Because `sf::st_intersect()` evaluates row wise, this minimizes the number of 
evaluations since the number of polygon features in barrier is always going to be much
much smaller than the number of edges.

With our network created, we need to identify the nearest node on the network for 
the start and end points of each identified segment. The `prt_nearestnode()` function
relies on the `{nabor}` package for fast determination of the nearest nodes.

```{r}
vis_graph <- prt_visgraph(land_barrier, buffer = 150)
segs_tbl <- segs_tbl %>% prt_nearestnode(vis_graph)
```

Notice, we pass `segs_tbl` as input to `prt_nearestnode()`. Each of these steps simply 
modifies the data stored in the original tibble output from `get_barrier_segments()`. We
now have all the information we need to calculate our shortest path routes around the 
barrier for each segment.

```{r}
segs_tbl <- segs_tbl %>% prt_shortpath(vis_graph)

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(segs_tbl$geometry, color = "deepskyblue3") +
  theme_void()

```

This process of 

1. `get_barrier_segments()`
2. `prt_nearestnode()`
3. `prt_shortpath()`

are likely to be the same across most workflows. So, the `prt_reroute()` function exists
to simply the process for the user. In short, this function calls the three functions
above along with some error handling and data checking. In the example code below
`prt_reroute()` is followed up with a call to `prt_update_points()`. For even more
efficient coding, the `%>%` function could be used to pipe the result from `prt_reroute()`
into `prt_update_points()`. The result will be a version of `track_pts` with the re-routed
geometry updated in place. Users may want to construct their own function for updating
points while also tracking which rows were updated and adjusting other column values
as needed.

```{r}
track_pts_fix <- prt_reroute(track_pts, land_barrier, vis_graph)

track_pts_fix <- prt_update_points(track_pts_fix, track_pts)
```


## Did it work?

A final plot of our corrected path. The green dot is the start and the red dot then end.
The blue line is the shortest path fix and the black line is the new path. Red line
represents the original path.

If everything is working correctly, there should not be any black lines crossing land.

```{r}
track_pts <- track_pts %>% st_cast('LINESTRING')
track_line_fixed <- track_pts_fix %>% summarise(do_union = FALSE) %>% st_cast('LINESTRING')

ggplot() + 
  ggspatial::annotation_spatial(land_barrier, fill = "cornsilk3", size = 0) +
  ggspatial::layer_spatial(track_pts, color = "red3") +
  ggspatial::layer_spatial(segs_tbl$geometry, color = "deepskyblue3", size = 2) +
  ggspatial::layer_spatial(track_line_fixed) +
  ggspatial::layer_spatial(l_pts[1,], color = "darkgreen", size = 4) +
  ggspatial::layer_spatial(l_pts[15,], color = "darkred", size = 4) +
  theme_void()
```
